---
title: "Working with Groups"
author: "N. Frerebeau"
date: "`r Sys.Date()`"
output:
  markdown::html_format:
    options:
      toc: true
      number_sections: true
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{Working with Groups}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
## Install extra packages (if needed)
# install.packages("folio")

library(nexus)
```

Provenance studies typically rely on two approaches, which can be used together:

* Identification of groups among the artifacts being studied, based on mineralogical or geochemical criteria (*clustering*).
* Comparison with so-called reference groups, i.e. known geological sources or archaeological contexts (*classification*).

When coercing a `data.frame` to a `CompositionMatrix` object, **nexus** allows to specify whether an observation belongs to a specific group (or not):

```{r}
## Data from Wood and Liu 2023
data("bronze", package = "folio")

## Use the third column (dynasties) for grouping
coda <- as_composition(bronze, parts = 4:11, groups = 3)
```

`groups(x)` and `groups(x) <- value` allow to retrieve or set groups of an existing `CompositionMatrix`. Missing values (`NA`) or empty strings can be used to specify that a sample does not belong to any group.

Once groups have been defined, they can be used by further methods (e.g. plotting). 
Note that for better readability, you can `select` only some of the parts (e.g. major elements):

```{r barplot, fig.width=7, fig.height=7, out.width='100%'}
## Select major elements
major <- coda[, is_element_major(coda)]

## Compositional bar plot
barplot(major, order_rows = "Cu", space = 0)
```

```{r mean, eval=FALSE}
## Compositional mean by artefact
coda <- condense(coda, by = list(bronze$dynasty, bronze$reference))
```

# Multivariate Analysis
## Log-Ratio Analysis

```{r pca, fig.width=7, fig.height=7, out.width='50%', fig.show='hold'}
## CLR
clr <- transform_clr(coda, weights = TRUE)

## PCA
lra <- pca(clr)

## Visualize results
viz_individuals(lra, color = c("#004488", "#DDAA33", "#BB5566"))
viz_hull(x = lra, border = c("#004488", "#DDAA33", "#BB5566"))

viz_variables(lra)
```

# References

Aitchison, J. (1986). *The Statistical Analysis of Compositional Data. Monographs on Statistics and Applied Probability*. Londres, UK ; New York, USA: Chapman and Hall.

Egozcue, J. J., Pawlowsky-Glahn, V., Mateu-Figueras, G. and Barceló-Vidal, C. (2003). Isometric Logratio Transformations for Compositional Data Analysis. *Mathematical Geology*, 35(3): 279-300. DOI: [10.1023/A:1023818214614](https://doi.org/10.1023/A:1023818214614).

Greenacre, M. (2021). Compositional Data Analysis. *Annual Review of Statistics and Its Application*, 8(1): 271-299. DOI: [10.1146/annurev-statistics-042720-124436](https://doi.org/10.1146/annurev-statistics-042720-124436).

Hron, K., Filzmoser, P., de Caritat, P., Fišerová, E. and Gardlo, A. (2017). Weighted Pivot Coordinates for Compositional Data and Their Application to Geochemical Mapping. *Mathematical Geosciences*, 49(6): 797-814. DOI : [10.1007/s11004-017-9684-z](https://doi.org/10.1007/s11004-017-9684-z).

Weigand, P. C., Harbottle, G. and Sayre, E. (1977). Turquoise Sources and Source Analysisis: Mesoamerica and the Southwestern U.S.A. In J. Ericson & T. K. Earle (Eds.), *Exchange Systems in Prehistory*, 15-34. New York, NY: Academic Press.
